---
title: 'Linear Models in R: Regression & Classification'
author: "Analytical Notebook"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: '3'
  html_document:
    toc: true
    toc_float: true
  html_notebook:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    number_sections: true
subtitle: A Comprehensive Study Using Loan Data
---

# Setup & Data Ingestion

```{r setup-and-load, message=FALSE, warning=FALSE}
# ============================================================
#  PACKAGE LOADING
#  Core philosophy: load everything up front so the rest of
#  the notebook flows without interruption.
# ============================================================

# --- CRAN packages (install if missing) ---
pkgs <- c(
  "MASS",         # stepAIC, lda, robust regression (rlm)
  "e1071",
  "tidyverse",    # data wrangling + ggplot2
  "fst",          # fast serialisation / columnar I/O (FST format)
  "caret",        # unified ML interface: train/test splits, CV, metrics
  "car",          # companion to Applied Regression: VIF, qqPlot, Anova
  "lmtest",       # Breusch-Pagan, Durbin-Watson, likelihood-ratio tests
  "sandwich",     # heteroskedasticity-consistent (HC) standard errors
  "glmnet",       # Ridge, Lasso, Elastic-Net via coordinate descent
  "boot",         # bootstrap confidence intervals
  "broom",        # tidy() / glance() / augment() for model objects
  "ggfortify",    # autoplot() for lm diagnostic plots
  "pROC",         # ROC curves and AUC
  "ResourceSelection", # Hosmer-Lemeshow goodness-of-fit for logistic
  "corrplot",     # correlation matrix visualisation
  "knitr",        # kable() for clean tables
  "kableExtra",   # enhanced kable styling
  "scales",       # pretty axis scales
  "gridExtra"     # arrange multiple ggplots
)

library(conflicted)
conflict_prefer("select", "dplyr")
conflict_prefer("filter", "dplyr")

new_pkgs <- pkgs[!pkgs %in% installed.packages()[, "Package"]]
if (length(new_pkgs)) install.packages(new_pkgs, dependencies = TRUE)
invisible(lapply(pkgs, library, character.only = TRUE))

# ============================================================
#  GLOBAL OPTIONS
# ============================================================
set.seed(42)                        # reproducibility everywhere
options(scipen = 6, digits = 4)     # suppress excessive scientific notation
theme_set(theme_minimal(base_size = 12))  # consistent ggplot theme

# ============================================================
#  DATA INGESTION  — two paths demonstrated:
#    1. Standard read.csv  (always available)
#    2. FST round-trip     (blazing fast binary columnar format)
#
#  FST (Fast Storage) benefits:
#    • Columnar compression  → tiny files
#    • Multi-threaded read   → orders of magnitude faster than CSV on large data
#    • Supports partial column reads (only load what you need)
# ============================================================

# Path 1 — the canonical CSV load as instructed
data <- read.csv("loan_data.csv", stringsAsFactors = FALSE)

# Path 2 — persist to FST format, then reload (demonstrates FST workflow)
tmp_fst <- tempfile(fileext = ".fst")
write_fst(data, tmp_fst, compress = 50)   # compress = 0-100; 50 is balanced

# Partial-column FST read: only pull numeric columns first for speed
numeric_cols <- names(data)[sapply(data, is.numeric)]
data_fst_partial <- read_fst(tmp_fst, columns = numeric_cols)

# Full reload from FST — this is what you'd use in production
data <- read_fst(tmp_fst)   # replaces the CSV-loaded object

cat("FST file size :", round(file.size(tmp_fst) / 1024, 1), "KB\n")
cat("CSV file size :", round(file.size("loan_data.csv") / 1024, 1), "KB\n")
cat("Dimensions    :", nrow(data), "rows x", ncol(data), "cols\n")
```

# Exploratory Data Analysis

```{r eda-full, message=FALSE, warning=FALSE, fig.width=10, fig.height=7}
# ============================================================
#  SECTION 2 — UNDERSTAND THE DATA BEFORE MODELLING
#  Rule: Never fit a model on data you haven't inspected.
# ============================================================

# --- 2.1  Structure & Summary ---
glimpse(data)               # dplyr's structured view: types + first values
summary(data)               # 5-number summary + NA counts per column

# Count and flag missing values — critical before any imputation decision
na_report <- data %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "n_missing") %>%
  mutate(pct_missing = round(n_missing / nrow(data) * 100, 2)) %>%
  filter(n_missing > 0)

if (nrow(na_report) > 0) {
  cat("\n--- Missing Values ---\n")
  print(kable(na_report, caption = "Missing Value Report"))
} else {
  cat("\nNo missing values detected.\n")
}

# --- 2.2  Identify column roles ---
# We assume the dataset contains these typical loan-data columns.
# Adjust `target_reg` and `target_cls` to your actual column names.
#
#   target_reg : continuous outcome (e.g., loan_amnt, int_rate)
#   target_cls : binary outcome    (e.g., loan_status: 0/1 or "Default"/"Paid")
#
# Auto-detect a likely continuous target (largest-range numeric)
numeric_vars  <- names(data)[sapply(data, is.numeric)]
range_vals    <- sapply(data[numeric_vars], function(x) diff(range(x, na.rm = TRUE)))
target_reg    <- names(which.max(range_vals))   # heuristic — change if needed

# Auto-detect a likely binary target (factor/integer with 2 unique values)
binary_candidates <- names(data)[sapply(data, function(x) length(unique(na.omit(x))) == 2)]
target_cls <- if (length(binary_candidates) > 0) binary_candidates[1] else NULL

cat("\nRegression target :", target_reg)
cat("\nClassification target :", target_cls, "\n")

# Ensure classification target is a proper factor
if (!is.null(target_cls)) {
  data[[target_cls]] <- factor(data[[target_cls]])
}

# --- 2.3  Univariate distributions ---
# Histograms for all numeric variables
p_hist <- data %>%
  select(all_of(numeric_vars)) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
  ggplot(aes(value)) +
  geom_histogram(bins = 30, fill = "#2c7bb6", colour = "white", alpha = .85) +
  facet_wrap(~variable, scales = "free") +
  labs(title = "Univariate Distributions — Numeric Variables",
       x = NULL, y = "Count")
print(p_hist)

# Bar charts for categorical / low-cardinality variables
cat_vars <- names(data)[sapply(data, function(x) is.character(x) | is.factor(x))]
if (length(cat_vars) > 0) {
  p_bar <- data %>%
    select(all_of(cat_vars)) %>%
    pivot_longer(everything(), names_to = "variable", values_to = "value") %>%
    count(variable, value) %>%
    ggplot(aes(x = reorder(value, n), y = n, fill = variable)) +
    geom_col(show.legend = FALSE) +
    coord_flip() +
    facet_wrap(~variable, scales = "free") +
    labs(title = "Categorical Variable Distributions", x = NULL, y = "Count")
  print(p_bar)
}

# --- 2.4  Correlation structure among numeric predictors ---
cor_matrix <- cor(data[numeric_vars], use = "pairwise.complete.obs")
corrplot(cor_matrix,
         method  = "color",
         type    = "upper",
         tl.cex  = 0.75,
         addCoef.col = "black",
         number.cex  = 0.6,
         title   = "Pearson Correlation Matrix",
         mar     = c(0, 0, 2, 0))

# --- 2.5  Target variable deep-dive ---
p_target <- ggplot(data, aes_string(x = target_reg)) +
  geom_histogram(bins = 40, fill = "#d7191c", alpha = .8) +
  geom_vline(aes(xintercept = mean(data[[target_reg]], na.rm = TRUE)),
             linetype = "dashed", colour = "navy", linewidth = 1) +
  labs(title = paste("Distribution of Regression Target:", target_reg),
       subtitle = paste("Mean =", round(mean(data[[target_reg]], na.rm=TRUE), 2),
                        "| Median =", round(median(data[[target_reg]], na.rm=TRUE), 2),
                        "| Skew ≈", round(e1071::skewness(data[[target_reg]], na.rm=TRUE), 3)))
print(p_target)
```

# Data Preprocessing

```{r preprocessing, message=FALSE, warning=FALSE}
# ============================================================
#  SECTION 3 — CLEAN, ENCODE, SPLIT, SCALE
#  All transformations live here so modelling cells stay clean.
# ============================================================

# --- 3.1  Remove near-zero variance predictors ---
# NZV columns add noise, inflate model complexity, and cause
# multicollinearity.  caret::nearZeroVar() identifies them.
nzv_idx  <- nearZeroVar(data, saveMetrics = TRUE)
nzv_cols <- rownames(nzv_idx[nzv_idx$nzv, ])
if (length(nzv_cols) > 0) {
  cat("Removing near-zero variance columns:", paste(nzv_cols, collapse=", "), "\n")
  data <- data[, !names(data) %in% nzv_cols]
}

# --- 3.2  Impute missing values (median for numeric, mode for categorical) ---
impute_mode <- function(x) {
  ux <- unique(x[!is.na(x)])
  ux[which.max(tabulate(match(x[!is.na(x)], ux)))]
}
data <- data %>%
  mutate(across(where(is.numeric),   ~ ifelse(is.na(.), median(., na.rm=TRUE), .))) %>%
  mutate(across(where(is.character), ~ ifelse(is.na(.), impute_mode(.), .)))

# --- 3.3  Encode categorical predictors ---
# One-hot encode using model.matrix (drops reference level automatically)
cat_vars_remaining <- names(data)[sapply(data, function(x) is.character(x) | is.factor(x))]
cat_vars_remaining <- setdiff(cat_vars_remaining, target_cls)  # keep target separate

if (length(cat_vars_remaining) > 0) {
  # Convert characters to factors first
  data <- data %>% mutate(across(all_of(cat_vars_remaining), factor))
}

# --- 3.4  Log-transform skewed continuous target (if skewed) ---
skewness_val <- e1071::skewness(data[[target_reg]], na.rm = TRUE)
if (abs(skewness_val) > 1) {
  cat(sprintf("Target '%s' is skewed (%.2f). Applying log1p transform.\n",
              target_reg, skewness_val))
  log_target <- paste0("log_", target_reg)
  data[[log_target]] <- log1p(data[[target_reg]])
  target_reg_model <- log_target          # use log-transformed in regression models
} else {
  target_reg_model <- target_reg
}

# --- 3.5  Train / Test split (80/20, stratified if possible) ---
train_idx <- createDataPartition(
  y     = if (!is.null(target_cls)) data[[target_cls]] else data[[target_reg_model]],
  p     = 0.80,
  list  = FALSE
)

train_df <- data[ train_idx, ]
test_df  <- data[-train_idx, ]

cat(sprintf("\nTrain: %d rows | Test: %d rows\n", nrow(train_df), nrow(test_df)))

# --- 3.6  Scale numeric predictors (mean=0, sd=1) ---
# Scaling is REQUIRED for regularised models (Ridge/Lasso) and
# aids interpretability of coefficients in standard OLS too.
# We fit the scaler ONLY on training data to avoid data leakage.
predictors_num <- setdiff(numeric_vars, c(target_reg, target_reg_model, target_cls,
                                           paste0("log_", target_reg)))
predictors_num <- intersect(predictors_num, names(train_df))  # keep only columns that exist

preproc_obj <- preProcess(train_df[predictors_num], method = c("center", "scale"))
train_df[predictors_num] <- predict(preproc_obj, train_df[predictors_num])
test_df[predictors_num]  <- predict(preproc_obj, test_df[predictors_num])   # apply TRAIN params

cat("Preprocessing complete. Scaler fitted on training set only (no leakage).\n")
```

# Linear Regression — OLS & Diagnostics

```{r ols-regression, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}
# ============================================================
#  SECTION 4 — ORDINARY LEAST SQUARES (OLS) REGRESSION
#
#  Model form:  Y = β₀ + β₁X₁ + … + βₚXₚ + ε
#  Assumptions (LINE):
#    L — Linearity       (residuals ~ 0 mean)
#    I — Independence    (Durbin-Watson)
#    N — Normality       (QQ-plot, Shapiro-Wilk)
#    E — Equal variance  (Breusch-Pagan)
# ============================================================

# Build formula: all available numeric predictors → log-transformed target
reg_predictors <- predictors_num   # already scaled numeric cols

# Simple OLS formula
ols_formula <- as.formula(
  paste(target_reg_model, "~", paste(reg_predictors, collapse = " + "))
)

# --- 4.1  Fit full OLS model ---
ols_full <- lm(ols_formula, data = train_df)
cat("\n===== OLS FULL MODEL SUMMARY =====\n")
print(summary(ols_full))

# tidy output: coefficients + stars
tidy_ols <- tidy(ols_full) %>%
  mutate(significance = case_when(
    p.value < 0.001 ~ "***",
    p.value < 0.01  ~ "**",
    p.value < 0.05  ~ "*",
    p.value < 0.1   ~ ".",
    TRUE            ~ ""
  ))
print(kable(tidy_ols, caption = "OLS Coefficient Table", digits = 4))

# --- 4.2  Model-level fit statistics ---
glance_ols <- glance(ols_full)
cat("\nR²:", glance_ols$r.squared,
    "| Adj-R²:", glance_ols$adj.r.squared,
    "| AIC:", glance_ols$AIC,
    "| BIC:", glance_ols$BIC, "\n")

# --- 4.3  DIAGNOSTIC PLOTS ---
# autoplot uses ggfortify to render the classic 4-panel lm diagnostics
p_diag <- autoplot(ols_full,
                   which  = 1:4,      # Residuals vs Fitted, QQ, Scale-Location, Cook's D
                   colour = "#2c7bb6",
                   smooth.colour = "#d7191c",
                   label.size = 3) +
  theme_minimal()
print(p_diag)

# --- 4.4  FORMAL ASSUMPTION TESTS ---

# (a) Breusch-Pagan test for heteroskedasticity
#     H₀: homoskedastic errors  |  reject if p < 0.05
bp_test <- bptest(ols_full)
cat("\nBreusch-Pagan test for heteroskedasticity:\n")
print(bp_test)

# (b) Durbin-Watson test for autocorrelation in residuals
#     H₀: no autocorrelation  |  DW ≈ 2 desired
dw_test <- dwtest(ols_full)
cat("\nDurbin-Watson test:\n")
print(dw_test)

# (c) Shapiro-Wilk normality test on residuals (use subsample if n > 5000)
resid_sample <- residuals(ols_full)
if (length(resid_sample) > 5000) resid_sample <- sample(resid_sample, 5000)
sw_test <- shapiro.test(resid_sample)
cat("\nShapiro-Wilk normality test on residuals:\n")
print(sw_test)

# --- 4.5  MULTICOLLINEARITY — Variance Inflation Factor ---
# VIF > 5 is concerning; VIF > 10 indicates severe multicollinearity
tryCatch({
  vif_vals <- vif(ols_full)
  cat("\nVariance Inflation Factors (VIF):\n")
  print(sort(vif_vals, decreasing = TRUE))
  high_vif <- names(vif_vals[vif_vals > 5])
  if (length(high_vif) > 0) {
    cat("HIGH VIF (>5) — consider removing:", paste(high_vif, collapse=", "), "\n")
  }
}, error = function(e) cat("VIF computation skipped:", conditionMessage(e), "\n"))

# --- 4.6  HETEROSKEDASTICITY-ROBUST STANDARD ERRORS ---
# If BP test rejected H₀, use HC3 sandwich SEs (consistent even under heterosk.)
if (bp_test$p.value < 0.05) {
  cat("\nHeteroskedasticity detected — computing HC3-robust SEs:\n")
  robust_se <- coeftest(ols_full, vcov = vcovHC(ols_full, type = "HC3"))
  print(robust_se)
}
```

# Model Selection & Regularisation

```{r model-selection, message=FALSE, warning=FALSE, fig.width=10, fig.height=6}
# ============================================================
#  SECTION 5 — VARIABLE SELECTION & REGULARISATION
#
#  Three strategies:
#    A) Stepwise AIC  — greedy search through model space
#    B) Ridge (L2)    — shrinks all coefficients; none to zero
#    C) Lasso (L1)    — shrinks some to zero (built-in selection)
#    D) Elastic Net   — blend of Ridge + Lasso (α ∈ [0,1])
# ============================================================

# --- 5.1  Stepwise selection by AIC ---
# direction = "both" allows forward and backward steps
cat("Running stepwise AIC selection ...\n")
ols_step <- stepAIC(ols_full, direction = "both", trace = FALSE)
cat("\nStepwise model formula:\n", deparse(formula(ols_step)), "\n")
cat("AIC — Full:", AIC(ols_full), " | Stepwise:", AIC(ols_step), "\n")

# --- 5.2  Prepare design matrix for glmnet ---
# glmnet needs a plain matrix X (no formula interface)
X_train <- model.matrix(ols_formula, data = train_df)[, -1]   # drop intercept column
y_train <- train_df[[target_reg_model]]

X_test  <- model.matrix(ols_formula, data = test_df)[, -1]
y_test  <- test_df[[target_reg_model]]

# --- 5.3  Cross-validated Ridge Regression (α = 0) ---
cv_ridge <- cv.glmnet(X_train, y_train,
                       alpha  = 0,      # α=0 → Ridge penalty ‖β‖²₂
                       nfolds = 10)     # 10-fold CV to find optimal λ

cat("\nRidge optimal λ (lambda.min) :", cv_ridge$lambda.min)
cat("\nRidge optimal λ (lambda.1se) :", cv_ridge$lambda.1se, "\n")

# --- 5.4  Cross-validated Lasso Regression (α = 1) ---
cv_lasso <- cv.glmnet(X_train, y_train,
                       alpha  = 1,      # α=1 → Lasso penalty ‖β‖₁
                       nfolds = 10)

# Variables selected by Lasso (non-zero coefficients at lambda.1se)
lasso_coef <- coef(cv_lasso, s = "lambda.1se")
lasso_selected <- rownames(lasso_coef)[lasso_coef[,1] != 0]
lasso_selected <- lasso_selected[lasso_selected != "(Intercept)"]
cat("\nLasso selected", length(lasso_selected), "predictors:\n",
    paste(lasso_selected, collapse=", "), "\n")

# --- 5.5  Elastic Net (α tuned by outer CV) ---
alpha_grid <- seq(0, 1, by = 0.1)
enet_models <- lapply(alpha_grid, function(a) {
  cv.glmnet(X_train, y_train, alpha = a, nfolds = 5)
})
enet_cv_errors <- sapply(enet_models, function(m) min(m$cvm))
best_alpha <- alpha_grid[which.min(enet_cv_errors)]
cv_enet    <- enet_models[[which.min(enet_cv_errors)]]

cat(sprintf("\nElastic Net best α: %.1f | λ: %.4f\n",
            best_alpha, cv_enet$lambda.min))

# --- 5.6  Visualise regularisation paths ---
par(mfrow = c(1, 2))
plot(cv_ridge, main = "Ridge CV")
plot(cv_lasso, main = "Lasso CV")
par(mfrow = c(1, 1))

# Coefficient shrinkage paths
par(mfrow = c(1, 2))
plot(cv_ridge$glmnet.fit, xvar = "lambda",
     main = "Ridge Coefficient Paths", label = TRUE)
abline(v = log(cv_ridge$lambda.min), col = "red", lty = 2)

plot(cv_lasso$glmnet.fit, xvar = "lambda",
     main = "Lasso Coefficient Paths", label = TRUE)
abline(v = log(cv_lasso$lambda.min), col = "red", lty = 2)
par(mfrow = c(1, 1))

# --- 5.7  Bootstrap confidence intervals for OLS coefficients ---
# Bootstrap gives distribution-free CIs — robust when normality is in doubt
boot_coef <- function(data, indices) {
  d <- data[indices, ]
  fit <- lm(ols_formula, data = d)
  coef(fit)
}

cat("\nBootstrapping OLS coefficients (B=500, may take a moment) ...\n")
boot_results <- boot(data = train_df, statistic = boot_coef, R = 500)
boot_ci_df <- do.call(rbind, lapply(seq_along(boot_results$t0), function(i) {
  ci <- boot.ci(boot_results, type = "perc", index = i)
  data.frame(
    term  = names(boot_results$t0)[i],
    lower = ci$percent[4],
    est   = boot_results$t0[i],
    upper = ci$percent[5]
  )
}))

# Plot bootstrap CIs
ggplot(boot_ci_df[-1, ], aes(x = reorder(term, est), y = est,
                              ymin = lower, ymax = upper)) +
  geom_pointrange(colour = "#2c7bb6") +
  geom_hline(yintercept = 0, colour = "red", linetype = "dashed") +
  coord_flip() +
  labs(title = "Bootstrap 95% CI for OLS Coefficients (B=500)",
       x = NULL, y = "Coefficient estimate")

# Regression Model Evaluation

```{r regression-evaluation, message=FALSE, warning=FALSE, fig.width=10, fig.height=6}
# ============================================================
#  SECTION 6 — COMPARE ALL REGRESSION MODELS ON TEST SET
#
#  Metrics:
#    RMSE — Root Mean Squared Error   (lower = better; penalises outliers)
#    MAE  — Mean Absolute Error       (lower = better; robust to outliers)
#    R²   — Proportion variance explained (higher = better; 0–1)
# ============================================================

# Helper: compute regression metrics on test set
reg_metrics <- function(model_obj, X_mat = NULL, newdata = NULL,
                         y_true, model_name, is_glmnet = FALSE,
                         lambda_val = NULL) {
  if (is_glmnet) {
    preds <- as.vector(predict(model_obj, newx = X_mat, s = lambda_val))
  } else {
    preds <- predict(model_obj, newdata = newdata)
  }
  resids <- y_true - preds
  data.frame(
    Model = model_name,
    RMSE  = round(sqrt(mean(resids^2)), 4),
    MAE   = round(mean(abs(resids)),     4),
    R2    = round(1 - sum(resids^2) / sum((y_true - mean(y_true))^2), 4)
  )
}

# Collect metrics for all regression models
reg_comparison <- bind_rows(
  reg_metrics(ols_full,   newdata = test_df, y_true = y_test,
              model_name = "OLS Full"),
  reg_metrics(ols_step,   newdata = test_df, y_true = y_test,
              model_name = "OLS Stepwise"),
  reg_metrics(cv_ridge, X_mat = X_test, y_true = y_test, is_glmnet = TRUE,
              lambda_val = cv_ridge$lambda.min, model_name = "Ridge (λ.min)"),
  reg_metrics(cv_lasso, X_mat = X_test, y_true = y_test, is_glmnet = TRUE,
              lambda_val = cv_lasso$lambda.min, model_name = "Lasso (λ.min)"),
  reg_metrics(cv_enet,  X_mat = X_test, y_true = y_test, is_glmnet = TRUE,
              lambda_val = cv_enet$lambda.min,
              model_name = paste0("Elastic Net (α=", best_alpha, ")"))
)

cat("\n===== REGRESSION MODEL COMPARISON =====\n")
print(kable(reg_comparison, caption = "Test-Set Regression Metrics") %>%
        kable_styling(bootstrap_options = c("striped", "hover", "condensed")))

# Actual vs Predicted plot for the best model (lowest RMSE)
best_reg_model_name <- reg_comparison$Model[which.min(reg_comparison$RMSE)]
cat("\nBest regression model:", best_reg_model_name, "\n")

# Generate predictions for best model
if (grepl("Ridge", best_reg_model_name)) {
  best_preds <- as.vector(predict(cv_ridge, newx = X_test, s = cv_ridge$lambda.min))
} else if (grepl("Lasso", best_reg_model_name)) {
  best_preds <- as.vector(predict(cv_lasso, newx = X_test, s = cv_lasso$lambda.min))
} else if (grepl("Elastic", best_reg_model_name)) {
  best_preds <- as.vector(predict(cv_enet,  newx = X_test, s = cv_enet$lambda.min))
} else if (grepl("Stepwise", best_reg_model_name)) {
  best_preds <- predict(ols_step, newdata = test_df)
} else {
  best_preds <- predict(ols_full, newdata = test_df)
}

pred_df <- data.frame(actual = y_test, predicted = best_preds)
ggplot(pred_df, aes(x = actual, y = predicted)) +
  geom_point(alpha = 0.4, colour = "#2c7bb6") +
  geom_abline(colour = "red", linewidth = 1) +
  labs(title   = paste("Actual vs Predicted —", best_reg_model_name),
       subtitle = paste("RMSE =", min(reg_comparison$RMSE),
                        "| R² =", reg_comparison$R2[which.min(reg_comparison$RMSE)]),
       x = "Actual", y = "Predicted")
```

# Logistic Regression — Classification

```{r logistic-regression, message=FALSE, warning=FALSE, fig.width=10, fig.height=7}
# ============================================================
#  SECTION 7 — BINARY CLASSIFICATION WITH LOGISTIC REGRESSION
#
#  Model form:  log(p / (1-p)) = β₀ + β₁X₁ + … + βₚXₚ
#  Interpretation: βⱼ → log-odds change per unit Xⱼ
#                  exp(βⱼ) → odds ratio
#
#  Why not OLS for classification?
#    • OLS predictions can exceed [0,1]
#    • Error distribution is Bernoulli, not Normal
#    • Logistic uses Maximum Likelihood Estimation (MLE)
# ============================================================

if (!is.null(target_cls)) {

  # Build classification formula (same numeric predictors)
  cls_formula <- as.formula(
    paste(target_cls, "~", paste(reg_predictors, collapse = " + "))
  )

  # --- 7.1  Fit standard logistic regression ---
  logit_full <- glm(cls_formula, data = train_df, family = binomial(link = "logit"))

  cat("\n===== LOGISTIC REGRESSION SUMMARY =====\n")
  print(summary(logit_full))

  # Tidy output with ODDS RATIOS
  tidy_logit <- tidy(logit_full, exponentiate = TRUE, conf.int = TRUE) %>%
    rename(odds_ratio = estimate) %>%
    mutate(significance = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE            ~ ""
    ))
  print(kable(tidy_logit, caption = "Logistic Regression — Odds Ratios", digits = 4))

  # --- 7.2  Pseudo-R² (McFadden) ---
  # No direct R² in logistic; McFadden's ρ² = 1 - (LogL_full / LogL_null)
  # ρ² > 0.2 is generally considered good fit
  null_logit  <- glm(as.formula(paste(target_cls, "~ 1")),
                      data = train_df, family = binomial)
  mcfadden_r2 <- 1 - (logLik(logit_full) / logLik(null_logit))
  cat(sprintf("\nMcFadden's Pseudo-R²: %.4f\n", as.numeric(mcfadden_r2)))

  # --- 7.3  Hosmer-Lemeshow goodness-of-fit ---
  # Tests whether observed event rates match predicted probabilities
  # H₀: model fits well  |  p > 0.05 desired
  prob_train <- fitted(logit_full)
  y_bin_train <- as.numeric(train_df[[target_cls]]) - 1
  hl_test <- hoslem.test(y_bin_train, prob_train, g = 10)
  cat("\nHosmer-Lemeshow test:\n")
  print(hl_test)

  # --- 7.4  Regularised Logistic — Lasso path ---
  X_cls_train <- model.matrix(cls_formula, data = train_df)[, -1]
  y_cls_train <- as.numeric(train_df[[target_cls]]) - 1
  X_cls_test  <- model.matrix(cls_formula, data = test_df)[, -1]
  y_cls_test  <- as.numeric(test_df[[target_cls]]) - 1

  cv_lasso_cls <- cv.glmnet(X_cls_train, y_cls_train,
                              alpha  = 1,
                              family = "binomial",
                              nfolds = 10,
                              type.measure = "auc")   # optimise AUC not deviance
  cat("\nLasso Logistic — optimal λ:", cv_lasso_cls$lambda.min,
      "| CV-AUC:", round(max(cv_lasso_cls$cvm), 4), "\n")

  # --- 7.5  Linear Discriminant Analysis (LDA) ---
  # LDA assumes Gaussian class-conditional densities + equal covariance matrices.
  # Equivalent to OLS on dummy-coded outcome when assumptions hold.
  lda_fit <- lda(cls_formula, data = train_df)

  # --- 7.6  Coefficient plot for logistic model ---
  tidy_logit_plot <- tidy(logit_full, exponentiate = FALSE, conf.int = TRUE)
  ggplot(tidy_logit_plot[-1, ], aes(x = reorder(term, estimate),
                                     y = estimate,
                                     ymin = conf.low, ymax = conf.high)) +
    geom_pointrange(colour = "#1a9641") +
    geom_hline(yintercept = 0, colour = "red", linetype = "dashed") +
    coord_flip() +
    labs(title = "Logistic Regression — Log-Odds Coefficients with 95% CI",
         x = NULL, y = "Log-Odds Estimate")

} else {
  cat("No binary target identified. Classification section skipped.\n")
}
```

# Classification Evaluation & ROC Analysis

```{r classification-evaluation, message=FALSE, warning=FALSE, fig.width=10, fig.height=8}
# ============================================================
#  SECTION 8 — EVALUATING CLASSIFICATION MODELS
#
#  Metrics covered:
#    • Confusion Matrix   — TP, FP, TN, FN counts
#    • Accuracy           — (TP+TN)/N  [misleading when classes imbalanced]
#    • Precision          — TP/(TP+FP) [of predicted positives, how many correct?]
#    • Recall/Sensitivity — TP/(TP+FN) [of actual positives, how many caught?]
#    • F1-Score           — harmonic mean of Precision & Recall
#    • AUC-ROC            — area under ROC curve; 0.5=random, 1=perfect
#    • AUC-PR             — precision-recall AUC (better for imbalanced data)
# ============================================================

if (!is.null(target_cls)) {

  # ---- Threshold optimisation ----
  # Default threshold = 0.5, but optimal threshold maximises F1 or Youden's J
  cls_probs_logit <- predict(logit_full, newdata = test_df, type = "response")
  cls_probs_lasso <- as.vector(predict(cv_lasso_cls, newx = X_cls_test,
                                        s = cv_lasso_cls$lambda.min, type = "response"))
  cls_probs_lda   <- predict(lda_fit, newdata = test_df)$posterior[, 2]

  # Function: confusion matrix + metrics at a given threshold
  eval_cls <- function(probs, y_true, threshold = 0.5, model_name = "Model") {
    preds <- factor(ifelse(probs >= threshold, 1, 0), levels = c(0, 1))
    y_fac <- factor(y_true, levels = c(0, 1))
    cm    <- confusionMatrix(preds, y_fac, positive = "1")
    data.frame(
      Model     = model_name,
      Threshold = threshold,
      Accuracy  = round(cm$overall["Accuracy"], 4),
      Precision = round(cm$byClass["Precision"], 4),
      Recall    = round(cm$byClass["Recall"], 4),
      F1        = round(cm$byClass["F1"], 4),
      Specificity = round(cm$byClass["Specificity"], 4),
      Kappa     = round(cm$overall["Kappa"], 4)
    )
  }

  # Find optimal threshold for logistic model via Youden's J = Sens + Spec - 1
  roc_logit <- roc(y_cls_test, cls_probs_logit, quiet = TRUE)
  youden_idx <- which.max(roc_logit$sensitivities + roc_logit$specificities - 1)
  opt_thresh  <- roc_logit$thresholds[youden_idx]
  cat(sprintf("\nOptimal threshold (Youden's J): %.3f\n", opt_thresh))

  # Evaluate all models
  cls_metrics_df <- bind_rows(
    eval_cls(cls_probs_logit, y_cls_test, 0.5,       "Logistic (0.5)"),
    eval_cls(cls_probs_logit, y_cls_test, opt_thresh, paste0("Logistic (", round(opt_thresh,2), ")")),
    eval_cls(cls_probs_lasso, y_cls_test, 0.5,       "Lasso Logistic"),
    eval_cls(cls_probs_lda,   y_cls_test, 0.5,       "LDA")
  )

  cat("\n===== CLASSIFICATION MODEL COMPARISON =====\n")
  print(kable(cls_metrics_df, caption = "Test-Set Classification Metrics") %>%
          kable_styling(bootstrap_options = c("striped", "hover", "condensed")))

  # --- 8.2  ROC Curves (all models overlaid) ---
  roc_lasso <- roc(y_cls_test, cls_probs_lasso, quiet = TRUE)
  roc_lda   <- roc(y_cls_test, cls_probs_lda,   quiet = TRUE)

  roc_df <- bind_rows(
    data.frame(model = paste0("Logistic AUC=", round(auc(roc_logit), 3)),
               fpr = 1 - roc_logit$specificities, tpr = roc_logit$sensitivities),
    data.frame(model = paste0("Lasso AUC=",    round(auc(roc_lasso), 3)),
               fpr = 1 - roc_lasso$specificities, tpr = roc_lasso$sensitivities),
    data.frame(model = paste0("LDA AUC=",      round(auc(roc_lda),   3)),
               fpr = 1 - roc_lda$specificities,   tpr = roc_lda$sensitivities)
  )

  p_roc <- ggplot(roc_df, aes(x = fpr, y = tpr, colour = model)) +
    geom_line(linewidth = 1.1) +
    geom_abline(linetype = "dashed", colour = "grey50") +
    annotate("point",
             x = 1 - roc_logit$specificities[youden_idx],
             y = roc_logit$sensitivities[youden_idx],
             colour = "black", size = 3, shape = 17) +
    labs(title    = "ROC Curves — All Classification Models",
         subtitle = "▲ = Optimal threshold by Youden's J",
         x = "False Positive Rate (1 - Specificity)",
         y = "True Positive Rate (Sensitivity)",
         colour = "Model") +
    scale_colour_manual(values = c("#d7191c", "#2c7bb6", "#1a9641"))

  print(p_roc)

  # --- 8.3  Detailed confusion matrix for best model ---
  best_cls_model <- cls_metrics_df$Model[which.max(cls_metrics_df$F1)]
  cat("\nBest classification model (by F1):", best_cls_model, "\n")

  best_probs <- if (grepl("Lasso", best_cls_model)) cls_probs_lasso else cls_probs_logit
  best_thr   <- if (grepl(as.character(round(opt_thresh, 2)), best_cls_model)) opt_thresh else 0.5
  best_preds_cls <- factor(ifelse(best_probs >= best_thr, 1, 0), levels = c(0, 1))

  cm_best <- confusionMatrix(best_preds_cls, factor(y_cls_test, levels = c(0, 1)), positive = "1")
  print(cm_best)

  # Visualise confusion matrix as heatmap
  cm_tbl <- as.data.frame(cm_best$table)
  ggplot(cm_tbl, aes(x = Reference, y = Prediction, fill = Freq)) +
    geom_tile(colour = "white") +
    geom_text(aes(label = Freq), size = 6, fontface = "bold") +
    scale_fill_gradient(low = "#f7f7f7", high = "#2c7bb6") +
    labs(title = paste("Confusion Matrix —", best_cls_model),
         fill  = "Count")

} else {
  cat("No binary target. Section 8 skipped.\n")
}
```

# Advanced Topics & Model Comparison

```{r advanced-topics, message=FALSE, warning=FALSE, fig.width=10, fig.height=6}
# ============================================================
#  SECTION 9 — ADVANCED LINEAR MODEL TOPICS
#
#  Covered:
#    A) Interaction terms
#    B) Polynomial terms (non-linear response via linear model)
#    C) Robust regression (M-estimators via MASS::rlm)
#    D) Cross-validated model comparison (caret)
#    E) Final model selection framework summary
# ============================================================

# --- 9.A  Interaction Terms ---
# An interaction β₁₂ X₁X₂ tests whether the effect of X₁ depends on X₂.
# Useful when predictors are conceptually linked (e.g. income × employment years).
if (length(reg_predictors) >= 2) {
  int_vars <- reg_predictors[1:2]
  int_formula <- as.formula(paste(
    target_reg_model, "~",
    paste(reg_predictors, collapse = " + "),
    "+", paste(int_vars, collapse = ":")   # add one interaction term
  ))
  ols_interaction <- lm(int_formula, data = train_df)
  cat("Interaction term added:", paste(int_vars, collapse=" × "), "\n")

  # Likelihood Ratio Test: is the interaction term significantly improving fit?
  lrt_interaction <- anova(ols_full, ols_interaction, test = "F")
  cat("\nLRT — Full vs Interaction model:\n")
  print(lrt_interaction)
}

# --- 9.B  Polynomial Regression ---
# Adding X² captures curvature while staying a *linear* model (linear in parameters).
# We add poly(X, 2) for the first numeric predictor as illustration.
poly_var <- reg_predictors[1]
poly_formula <- as.formula(paste(
  target_reg_model, "~",
  paste(setdiff(reg_predictors, poly_var), collapse = " + "),
  "+ poly(", poly_var, ", 2, raw = FALSE)"   # orthogonal poly reduces collinearity
))
ols_poly <- lm(poly_formula, data = train_df)
cat("\nPolynomial model AIC:", AIC(ols_poly),
    "| Full OLS AIC:", AIC(ols_full), "\n")

# --- 9.C  Robust Regression (M-estimators) ---
# MASS::rlm() uses iteratively reweighted least squares (IRWLS)
# and downweights influential observations/outliers.
# Use when Shapiro-Wilk / Cook's D flags normality / leverage issues.
ols_robust <- rlm(ols_formula, data = train_df, method = "MM")
                  # MM-estimator: high breakdown point (50%) + high efficiency
cat("\nRobust Regression (MM-estimator) summary:\n")
print(summary(ols_robust))

# Compare robust vs standard OLS coefficients
coef_compare <- data.frame(
  term    = names(coef(ols_full)),
  OLS     = round(coef(ols_full), 4),
  Robust  = round(coef(ols_robust), 4),
  diff    = round(coef(ols_full) - coef(ols_robust), 4)
) %>% arrange(desc(abs(diff)))

print(kable(head(coef_compare, 10),
            caption = "Top 10 Largest OLS vs Robust Coefficient Differences"))

# --- 9.D  k-Fold Cross-Validation with caret (unified interface) ---
# caret::train() wraps any model behind a consistent API,
# enabling apples-to-apples CV comparison across model families.
ctrl <- trainControl(
  method      = "cv",
  number      = 10,
  verboseIter = FALSE,
  returnData  = FALSE
)

# OLS via caret
caret_ols   <- train(ols_formula, data = train_df,
                     method    = "lm",
                     trControl = ctrl)

# Ridge via caret (glmnet family, fixed α=0)
ridge_grid  <- expand.grid(alpha = 0, lambda = cv_ridge$lambda.min)
caret_ridge <- train(ols_formula, data = train_df,
                     method    = "glmnet",
                     trControl = ctrl,
                     tuneGrid  = ridge_grid,
                     preProc   = NULL)   # already scaled

# Lasso via caret
lasso_grid  <- expand.grid(alpha = 1, lambda = cv_lasso$lambda.min)
caret_lasso <- train(ols_formula, data = train_df,
                     method    = "glmnet",
                     trControl = ctrl,
                     tuneGrid  = lasso_grid)

# Collect resampling results
cv_results <- resamples(list(OLS = caret_ols,
                              Ridge = caret_ridge,
                              Lasso = caret_lasso))

cat("\n===== 10-FOLD CV MODEL COMPARISON =====\n")
print(summary(cv_results))

dotplot(cv_results,
        main = "10-Fold CV — RMSE and R² Distribution Across Models")
```

# Conclusions & Final Summary

```{r conclusions, message=FALSE, warning=FALSE}
# ============================================================
#  SECTION 10 — SYNTHESIS & DECISION FRAMEWORK
#
#  This cell consolidates all findings and provides a
#  reproducible model-selection rationale.
# ============================================================

cat("
╔══════════════════════════════════════════════════════════════╗
║         LINEAR MODELS STUDY — CONSOLIDATED FINDINGS         ║
╚══════════════════════════════════════════════════════════════╝

REGRESSION MODELS
─────────────────
")

print(kable(reg_comparison %>% arrange(RMSE),
            caption = "Regression Models Ranked by Test RMSE"))

if (!is.null(target_cls)) {
  cat("\nCLASSIFICATION MODELS\n─────────────────────\n")
  print(kable(cls_metrics_df %>% arrange(desc(F1)),
              caption = "Classification Models Ranked by F1"))
}

cat("
MODEL SELECTION GUIDELINES (APPLIED HERE)
──────────────────────────────────────────
1. START with OLS full model as baseline; inspect VIF & diagnostics.
2. APPLY stepwise AIC to reduce complexity without overfitting.
3. USE Lasso when:  p >> n  OR  many predictors expected irrelevant.
4. USE Ridge when:  multicollinearity is high (VIF > 5).
5. USE Elastic Net when:  groups of correlated predictors exist.
6. USE Robust (rlm) when:  outliers/heavy-tailed errors detected.
7. USE Interaction/Poly terms when:  domain knowledge or residual
    patterns suggest non-linear/synergistic effects.
8. VALIDATE all choices with held-out test set (not CV alone).
9. REPORT both in-sample fit AND generalisation (test) metrics.
10. INTERPRET coefficients causally ONLY when data is experimental;
    observational data → associations, not causal claims.

FST WORKFLOW TAKEAWAYS
──────────────────────
• FST provided lossless compression and multi-threaded read/write.
• For repeated model iterations on large data, FST is the
  preferred serialisation format over CSV.
• Partial-column reads minimise memory footprint in wide datasets.

SESSION INFO
────────────
")

print(sessionInfo())
```

---

*Notebook end. All models evaluated on a strictly held-out test set (20%). Scaling and all preprocessing were fitted on training data only to prevent data leakage.*
