---
title: "Loan Data Analysis & Modeling"
subtitle: "A Teaching Guide for R Beginners in RStudio"
author: "Valerie Jerono"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
# This block runs silently when you knit the document.
# include=FALSE means it won't appear in the output.
knitr::opts_chunk$set(
  echo    = TRUE,   # show code in output
  message = FALSE,  # hide package messages
  warning = FALSE   # hide warnings
)
```

---

> **Before you start:** Open this file in RStudio. You can run each code chunk  
> by clicking the green ▶ button on its top-right, or press **Ctrl+Shift+Enter**  
> (Windows/Linux) / **Cmd+Shift+Enter** (Mac). To produce the full HTML report,  
> click **Knit** in the toolbar at the top.

---

## 1 · Setup: Installing & Loading Packages

> **📘 What is a package?**  
> R's base installation is powerful, but the R community has written thousands of  
> *packages* (add-ons) that extend it. You install a package once with  
> `install.packages()`, then load it each session with `library()`.  
>  
> Think of it like: installing an app on your phone (once) vs. opening it (every time).

```{r install-packages, eval=FALSE}
# Run this block ONCE to install all required packages.
# After installation, you never need to run install.packages() again
# for these packages on your computer.

install.packages(c(
  "tidyverse",  # Collection of data science tools (ggplot2, dplyr, tidyr, etc.)
  "caret",      # Classification And REgression Training — machine learning toolkit
  "corrplot",   # Visualising correlation matrices
  "car",        # Companion to Applied Regression (includes VIF function)
  "MASS",       # Modern Applied Statistics (includes stepAIC)
  "ROCR"        # ROC curves for classifier evaluation
))
```

```{r load-libraries}
# Load packages every time you open a new R session.
# library() makes the package's functions available in your environment.

library(tidyverse)  # loads ggplot2, dplyr, tidyr, readr, and more
library(caret)
library(corrplot)
library(car)
library(MASS)
library(ROCR)
library(dplyr)      # loaded by tidyverse, but we call it explicitly for clarity
```

> **📘 Why does tidyverse load so many things?**  
> `tidyverse` is a "meta-package" — a curated collection. When you run  
> `library(tidyverse)`, it silently loads ~8 packages at once, saving you time.

---

## 2 · Load & Explore the Dataset

> **📘 Assigning variables with `<-`**  
> In R, the assignment operator is `<-` (not `=`). It means "store the result  
> on the right into the name on the left". You can read `data <- read.csv(...)` as  
> *"data gets the CSV file"*. Keyboard shortcut: **Alt + –** (Windows) / **Option + –** (Mac).

```{r load-data}
# Read the CSV file into a data frame called 'data'.
# A data frame is R's version of a spreadsheet/table — rows are observations,
# columns are variables.

# ⚠️  Update this path to wherever your loan_data.csv is saved on YOUR computer.
# Example on Windows: "C:/Users/YourName/Downloads/loan_data.csv"
# Example on Mac:     "/Users/YourName/Downloads/loan_data.csv"

data <- read.csv("loan_data.csv")

# head() shows the first 6 rows — a quick peek at the structure
head(data)
```

```{r summarise-data}
# summary() gives descriptive statistics for every column:
#   - For numeric columns: min, max, mean, median, quartiles
#   - For character/factor columns: frequency counts

summary(data)
```

> **📘 Reading the summary output**  
> - **Min / Max**: the smallest and largest values in that column  
> - **Mean**: the average (sum divided by count)  
> - **Median**: the middle value when sorted — less sensitive to outliers than mean  
> - **1st Qu. / 3rd Qu.**: the 25th and 75th percentile values  
>  
> When Mean >> Median, the data is *right-skewed* (a few very large values pull  
> the mean up). This is common for income data, which we will address later.

---

## 3 · Data Cleaning

### 3.1 Check for Missing Values

> **📘 `colSums(is.na(...))`**  
> - `is.na(data)` produces a TRUE/FALSE table — TRUE wherever a value is missing  
> - `colSums(...)` sums each column (TRUE = 1, FALSE = 0)  
> - Result: count of missing values per column

```{r missing-values}
# Count missing values in each column.
# A result of 0 for every column means the data is complete — no gaps to fill.
colSums(is.na(data))
```

### 3.2 Convert Categorical Columns to Factors

> **📘 What is a `factor`?**  
> R has a special data type for *categorical* variables called a **factor**.  
> A factor stores distinct categories (called *levels*) and encodes them  
> internally as integers. This matters for modelling — when R sees a factor,  
> it knows to create *dummy variables* automatically (more on this soon).  
>  
> Without conversion, R would treat a column like `person_gender` as plain text  
> and most modelling functions would either error or ignore it.

```{r convert-factors}
# Convert each categorical column from character to factor.
# Syntax: data$column_name  accesses a single column from a data frame.

data$person_gender              <- as.factor(data$person_gender)
data$person_education           <- as.factor(data$person_education)
data$person_home_ownership      <- as.factor(data$person_home_ownership)
data$loan_intent                <- as.factor(data$loan_intent)
data$previous_loan_defaults_on_file <- as.factor(data$previous_loan_defaults_on_file)

# loan_status is our classification target: 0 = did not default, 1 = defaulted
data$loan_status                <- as.factor(data$loan_status)

# Verify the changes — factor columns now show levels instead of raw text
summary(data)
```

---

## 4 · Exploratory Data Analysis (EDA)

### 4.1 Correlation Matrix

> **📘 What is correlation?**  
> Correlation measures the *linear* relationship between two numeric variables.  
> - **+1**: perfect positive relationship (as X increases, Y increases)  
> - **-1**: perfect negative relationship (as X increases, Y decreases)  
> - **0**: no linear relationship  
>  
> **Why does it matter?** Highly correlated *predictor* variables (correlation > 0.8)  
> can cause *multicollinearity* in regression models — we will check for this later.

```{r correlation-matrix, fig.width=8, fig.height=7}
# Step 1: Keep only numeric columns — cor() can't handle factors or text.
# The pipe operator %>% means "take the result and pass it to the next function".
# dplyr::select() specifies we want select() from dplyr (avoids name conflicts).
# where(is.numeric) is a helper that picks columns satisfying is.numeric.

numeric_data <- data %>% dplyr::select(where(is.numeric))

# Step 2: Compute the correlation matrix
cor_matrix <- cor(numeric_data)

# Step 3: Visualise with corrplot
# method = "color" fills each cell with a colour proportional to the correlation.
# Blue = positive correlation, Red = negative correlation.
corrplot(cor_matrix, method = "color",
         tl.cex = 0.8,    # text label size
         tl.col = "black") # text label colour
```

> **🔍 What to look for:**  
> - `person_age` ↔ `person_emp_exp`: very high correlation (~0.82) — these variables  
>   measure similar things. This will show up as *multicollinearity* in the model.  
> - `loan_amnt` ↔ `loan_percent_income`: moderate positive correlation (~0.64).

### 4.2 Income Distribution

> **📘 `hist()` — Histogram**  
> A histogram splits a numeric variable into equal-width *bins* and counts how  
> many observations fall into each bin. It reveals the *shape* of the distribution.

```{r income-histogram, fig.width=7, fig.height=5}
hist(data$person_income,
     breaks = 30,                  # number of bins
     main   = "Income Distribution",
     xlab   = "Annual Income",
     col    = "#7c6fff",
     border = "white")

# Add a vertical line at the mean and median to see the skew
abline(v = mean(data$person_income),   col = "red",   lwd = 2, lty = 2)
abline(v = median(data$person_income), col = "green", lwd = 2, lty = 2)
legend("topright",
       legend = c("Mean", "Median"),
       col    = c("red", "green"),
       lty    = 2, lwd = 2)
```

> **🔍 Interpretation:**  
> The distribution is **right-skewed** — most people earn low-to-moderate income,  
> with a long tail of high earners. The mean is pulled right of the median.  
>  
> **Why this matters:** Linear regression assumes residuals are normally distributed.  
> A skewed *target variable* often violates this. The fix: **log transformation**.

---

## 5 · Linear Regression — Predicting Income

> **📘 What is linear regression?**  
> Linear regression fits a straight-line relationship between a target variable Y  
> (what we want to predict) and one or more predictor variables X:  
>  
> `Y = β₀ + β₁X₁ + β₂X₂ + ... + ε`  
>  
> The model *learns* the β coefficients (slopes) that minimise prediction error.  
> In R, you describe this relationship with a **formula**: `target ~ predictors`.  
> Using `.` as the predictor means "use all other columns".

### 5.1 Train/Test Split

> **📘 Why split the data?**  
> We train the model on one portion of data and evaluate it on an *unseen* portion.  
> This tests whether the model *generalises* rather than just memorising training data  
> (a problem called *overfitting*).

```{r train-test-split}
set.seed(123)
# set.seed() ensures randomness is reproducible — run this code on any computer
# and you'll get the same split. The number 123 is arbitrary; any integer works.

# createDataPartition() from caret splits indices proportionally.
# p = 0.8 means 80% goes to training.
# list = FALSE returns a plain vector of row numbers, not a list.
train_index <- createDataPartition(data$person_income, p = 0.8, list = FALSE)

# Subset the data using the indices:
# data[train_index, ]  — rows in train_index,  all columns
# data[-train_index, ] — rows NOT in train_index (the minus sign = "exclude")
train <- data[train_index, ]
test  <- data[-train_index, ]

cat("Training rows:", nrow(train), "\n")
cat("Test rows    :", nrow(test),  "\n")
```

### 5.2 Fit Baseline Linear Model

```{r baseline-linear-model}
# lm() = linear model
# Formula: person_income ~ .  means: predict income using ALL other columns
# data = train: fit on training data only

lm_model <- lm(person_income ~ ., data = train)

summary(lm_model)
```

> **📘 Reading the `summary()` output:**  
> - **Coefficients table**: each row is one predictor. The `Estimate` column is  
>   the β coefficient — how much Y changes for a 1-unit increase in that X.  
>   `***` stars mean the predictor is statistically significant (p < 0.001).  
> - **R-squared**: proportion of variance in Y explained by the model.  
>   R² = 0.29 means the model explains **29%** of income variation — not great.  
> - **Residual Std. Error**: average prediction error in the same units as Y (income).

### 5.3 Check Regression Assumptions

> **📘 The 4 plots from `plot(model)`:**  
> 1. **Residuals vs Fitted** — checks linearity. Points should be randomly scattered  
>    around the horizontal line (no pattern = good).  
> 2. **Normal Q-Q** — checks normality of residuals. Points should follow the diagonal  
>    line. Deviations at the tails indicate non-normality.  
> 3. **Scale-Location** — checks homoscedasticity (equal variance). A flat line is good.  
> 4. **Residuals vs Leverage** — identifies influential outliers.

```{r check-assumptions, fig.width=8, fig.height=8}
# par(mfrow = c(2,2)) divides the plotting window into a 2×2 grid.
# The next 4 plots will fill each panel in order.
par(mfrow = c(2, 2))
plot(lm_model)
```

```{r vif-check}
# VIF = Variance Inflation Factor — measures multicollinearity.
# A VIF > 10 is a sign of serious multicollinearity.
# When two predictors are highly correlated, the model can't reliably
# estimate their individual effects.

vif(lm_model)
```

> **🔍 Findings:**  
> - `person_age` (GVIF ≈ 14.1) and `person_emp_exp` (GVIF ≈ 11.4) are highly  
>   correlated — the model struggles to separate their effects.  
> - The Q-Q plot shows non-normal residuals — a consequence of skewed income.  
> - **Solution**: log-transform the target variable.

---

## 6 · Transformations

### 6.1 Apply Log Transform

> **📘 Why log-transform?**  
> `log()` compresses large values and expands small ones, turning a right-skewed  
> distribution into something closer to a normal distribution.  
> This satisfies regression's normality assumption and often dramatically improves R².  
>  
> **Trade-off**: predictions will be on the *log scale*. You must `exp()` them  
> back to the original scale to interpret results in actual dollars.

```{r log-transform}
# Add a new column log_income to both train and test.
# log() in R is the natural logarithm (base e).
train$log_income <- log(train$person_income)
test$log_income  <- log(test$person_income)

# Quick check — the distribution should now look more symmetric
hist(train$log_income,
     breaks = 30,
     main   = "Log-Income Distribution",
     xlab   = "log(Annual Income)",
     col    = "darkgreen",
     border = "white")
```

### 6.2 Refit with Log Income

```{r refit-log-model}
# Formula:  log_income ~ . - person_income
# The minus sign ( - person_income ) REMOVES person_income from the predictors.
# We exclude it because log_income is derived from it — including both would
# be circular (and perfectly correlated).

lm_model_log <- lm(log_income ~ . - person_income, data = train)

summary(lm_model_log)
```

> **🔍 R² jumped from 0.29 → 0.77!**  
> The log transformation fixed the skewness and the model now explains  
> ~77% of income variance. This is a *huge* improvement from a simple fix.

### 6.3 Check Assumptions Again

```{r assumptions-log, fig.width=8, fig.height=8}
par(mfrow = c(2, 2))
plot(lm_model_log)

vif(lm_model_log)
```

---

## 7 · Model Evaluation

> **📘 Metrics for regression:**  
> - **RMSE** (Root Mean Squared Error): average error in original units. Penalises  
>   large errors more severely because errors are squared before averaging.  
> - **MAE** (Mean Absolute Error): average of absolute errors. More intuitive —  
>   "on average, our predictions are off by MAE dollars".  
> Lower is better for both.

```{r evaluate-linear-model}
# Step 1: Predict on the test set (data the model has never seen)
# newdata = test tells predict() to use the test data frame
pred <- predict(lm_model_log, newdata = test)

# Step 2: Back-transform from log scale to original income scale
# exp() is the inverse of log() — exp(log(x)) = x
pred_exp <- exp(pred)

# Step 3: Calculate error metrics
RMSE <- sqrt(mean((test$person_income - pred_exp)^2))
MAE  <- mean(abs(test$person_income - pred_exp))

cat("RMSE:", round(RMSE, 2), "\n")
cat("MAE :", round(MAE,  2), "\n")
```

> **🔍 Interpretation:**  
> MAE ≈ 16,591 means the model's predictions are off by about **$16,591** on  
> average. Whether that's acceptable depends on the business context.

---

## 8 · Stepwise Regression (Model Selection)

> **📘 What is stepwise regression?**  
> Stepwise regression automates variable selection. It iteratively adds or removes  
> predictors based on the **AIC** (Akaike Information Criterion) — a measure that  
> balances model fit against complexity. Lower AIC = better model.  
>  
> `direction = "both"` means the algorithm can both add and remove variables at  
> each step (also called *bidirectional elimination*).

```{r stepwise, results='hide'}
# stepAIC() is from the MASS package.
# It prints a step-by-step log as it runs — suppressed here with results='hide'
# to keep output clean. Remove that chunk option to see the full log.
step_model <- stepAIC(lm_model_log, direction = "both")
```

```{r stepwise-summary}
summary(step_model)
```

> **🔍 Result:**  
> Stepwise removed `credit_score`, `person_gender`, and `previous_loan_defaults_on_file`  
> without changing R². These variables were not adding meaningful predictive power.  
> The simpler model is preferred — fewer variables = easier to interpret and deploy.

---

## 9 · Classification Model — Predicting Loan Default

> **📘 What is logistic regression?**  
> Despite the name, logistic regression is a *classification* algorithm.  
> Instead of predicting a continuous number, it predicts the **probability** that  
> an observation belongs to a class (here: probability of loan default).  
>  
> We use `glm()` (Generalised Linear Model) with `family = "binomial"` — this  
> tells R we are modelling a binary outcome (0 or 1).

### 9.1 Fit the Logistic Model

```{r logistic-model}
# Target: loan_status (0 = no default, 1 = default)
# We exclude person_income (raw) because we already have log_income.
# family = "binomial" activates the logistic regression link function.

log_model <- glm(loan_status ~ . - person_income,
                 data   = train,
                 family = "binomial")

summary(log_model)
```

> **📘 Reading the logistic regression summary:**  
> - Coefficients are on the *log-odds* scale (not directly interpretable as probabilities).  
> - `exp(coef(log_model))` converts them to **odds ratios** — more interpretable.  
> - **AIC**: like in stepwise — lower is better. Use to compare models.  
> - **Null vs Residual deviance**: a large drop (null − residual) means the predictors  
>   are explaining a lot of variation in the outcome.

### 9.2 Predictions & Confusion Matrix

> **📘 The threshold (0.5):**  
> `predict(..., type = "response")` returns probabilities between 0 and 1.  
> We need to convert these to class labels using a threshold.  
> `ifelse(prob > 0.5, 1, 0)` means: "predict default if probability > 50%".  
> You can adjust this threshold to trade off sensitivity vs. specificity.

```{r confusion-matrix}
# type = "response" returns probabilities (not log-odds)
prob       <- predict(log_model, newdata = test, type = "response")

# Convert probabilities to binary predictions using 0.5 threshold
pred_class <- ifelse(prob > 0.5, 1, 0)

# confusionMatrix() from caret compares predictions vs. true labels
confusionMatrix(as.factor(pred_class), test$loan_status)
```

> **📘 Confusion Matrix explained:**  
>  
> |                  | Predicted: No Default | Predicted: Default |  
> |------------------|-----------------------|--------------------|  
> | **Actual: No Default** | True Negative (TN)  | False Positive (FP) |  
> | **Actual: Default**    | False Negative (FN)  | True Positive (TP)  |  
>  
> - **Accuracy** = (TP + TN) / total = 89.8% — correct predictions overall  
> - **Sensitivity** = TP / (TP + FN) = 93.8% — % of actual defaults correctly caught  
> - **Specificity** = TN / (TN + FP) = 76.0% — % of actual non-defaults correctly identified  
>  
> **High sensitivity** is often more important in lending — missing a default is  
> more costly than wrongly flagging a good borrower.

### 9.3 ROC Curve

> **📘 What is the ROC Curve?**  
> The **Receiver Operating Characteristic** curve plots sensitivity (true positive rate)  
> vs. 1 − specificity (false positive rate) at every possible threshold.  
>  
> A model that randomly guesses falls on the diagonal. A perfect model hits the  
> top-left corner. The **AUC** (Area Under the Curve) summarises performance:  
> - AUC = 0.5: no better than random  
> - AUC = 1.0: perfect  
> - AUC > 0.9: excellent

```{r roc-curve, fig.width=6, fig.height=6}
# prediction() and performance() are from the ROCR package.
# They create objects that know how to plot themselves.
pred_obj <- prediction(prob, test$loan_status)
perf     <- performance(pred_obj, "tpr", "fpr")

plot(perf,
     col     = "#7c6fff",
     lwd     = 2,
     main    = "ROC Curve — Loan Default Classifier",
     xlab    = "False Positive Rate (1 - Specificity)",
     ylab    = "True Positive Rate (Sensitivity)")

# Reference line: what a random classifier would look like
abline(a = 0, b = 1, col = "grey50", lty = 2)

# Calculate and display AUC
auc <- performance(pred_obj, "auc")@y.values[[1]]
legend("bottomright",
       legend = paste0("AUC = ", round(auc, 3)),
       col    = "#7c6fff", lty = 1, lwd = 2)
```

---

## 10 · Feature Engineering

> **📘 What is feature engineering?**  
> Feature engineering means creating *new* columns (features) from existing ones  
> that might be more informative for a model. Good features capture domain  
> knowledge — for example, "income relative to experience" captures *productivity*  
> in a way that raw income or experience alone do not.

```{r feature-engineering}
# Feature 1: income per year of experience
# Adding 1 prevents division by zero for people with 0 experience
data$income_per_experience <- data$person_income / (data$person_emp_exp + 1)

# Feature 2: loan amount as a fraction of credit score
# Higher ratio = potentially riskier loan relative to creditworthiness
data$loan_to_credit_ratio  <- data$loan_amnt / data$credit_score

# Inspect the new features
summary(data[, c("income_per_experience", "loan_to_credit_ratio")])
```

> **💡 Next steps:**  
> After engineering features, you would re-run the train/test split  
> (including the new columns) and refit your models to see if performance improves.

---

## 11 · Summary & Key Takeaways

```{r summary-table, echo=FALSE}
summary_df <- data.frame(
  Step     = c("Baseline Linear Regression",
               "Linear (after log transform)",
               "Stepwise Linear (AIC selection)",
               "Logistic Regression"),
  Metric   = c("R² = 0.293",
               "R² = 0.767",
               "R² = 0.767 (simpler model)",
               "Accuracy = 89.8%, AUC ≈ 0.93"),
  Key_Lesson = c("Skewed targets hurt model fit",
                 "Log transform: huge improvement",
                 "Fewer variables can be equally good",
                 "Classification: accuracy + AUC")
)

knitr::kable(summary_df,
             col.names = c("Model", "Performance", "Key Lesson"),
             caption   = "Model Comparison")
```

---

> **🎓 R Concepts Covered:**  
> `<-` assignment · `library()` · `read.csv()` · `head()` · `summary()` ·  
> `as.factor()` · `%>%` pipe · `dplyr::select()` · `cor()` · `corrplot()` ·  
> `hist()` · `set.seed()` · `createDataPartition()` · `lm()` · `glm()` ·  
> `par(mfrow)` · `plot(model)` · `vif()` · `log()` · `exp()` · `predict()` ·  
> `stepAIC()` · `ifelse()` · `confusionMatrix()` · `prediction()` · `performance()`

---

*Knit this document to HTML with the **Knit** button in RStudio to see the full formatted report.*
